Practical Optional Types for Clojure

### 2015-05-16 formal-model.tex
JVM ops are dynamically typed

Recall \egref{example:doexception}.
\begin{minted}{clojure}
...  (do (if (number? x) nil (throw (Exception.)))
         (inc x)) ...
\end{minted}

@@

\subsection{Java Interoperability and Type Hints}

\begin{figure}[t]
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots   \localhintsyntax{} \alt \lethintsyntax{} &\mbox{Expressions}\\
            &\alt& \reflectiveexpsyntax{} \\
     &\alt& \mininonreflectiveexpsyntax{}
    \\

@@

    \tatypesyntax{}\\
    \typehintenvsyntax{}\\
    \classtableallsyntax{}
  \end{altgrammar}
  $$
  \begin{mathpar}
    {\TNewStatic}

    {\TMethodStatic}

@@

    \BMethod{}
  \end{mathpar}
  \begin{mathpar}
    %\RAbs{}
%
%    \RNewElimRefl{}
%
%    \RMethodElimRefl{}
%
%    \RFieldElimRefl{}
%
%    \RLet{}
%

%    \RLetHint{}
  \end{mathpar}
  \caption{Java Interoperability Syntax, Typing and Operational Semantics}
  \label{main:figure:javatyping}
\end{figure}

In \secref{sec:overviewjavainterop} we discussed the role of type hints
to help eliminate reflective calls.
In this section, we introduce our model of Java and provide user-facing
syntax corresponding to Clojure's Java interoperability forms and type hinted forms.
Then we model the Clojure compiler's compile-time reflection resolution
algorithm.
To achieve this, first we define notation for
\emph{non-reflective} Java forms that 
unambiguously call a field, method or constructor.
Then we define a rewrite relation that uses
type hints to resolve reflection explicitly.
Finally we give typing rules that model how
Typed Clojure interacts with non-reflective calls.

@@

We extend the syntax in \figref{main:figure:javatyping} with
type hinted expressions, reflective and non-reflective Java field lookups and calls to
methods and constructors. 
We model the syntax after the `dot' special
form to prevent ambiguity---\clj{(.fld e)} is now \fieldexp{\fld{}}{\e{}},
\clj{(.mth e es*)} is $\methodexp{\mth{}}{\e{}}{\overrightarrow{es}}$
and \clj{(.class es*)} is $\newexp{\class{}}{\overrightarrow{es}}$.
The reflective expressions come without typing rules because Typed Clojure
only reasons about resolved reflection
(as demonstrated in \secref{sec:overviewjavainterop}).
%
The method call ${\methodstaticexp {\classhint{1}}
                             {\overrightarrow {\classhint{i}}}
                             {\classhint{2}}
                             {\mth{}} {\e{}} {\overrightarrow{\e{i}}}}$
is a non-reflective call to the \mth{} method on class {\class{1}}, 
with Java signature 
${\classhint{2}}\ \mth{}\ (\overrightarrow {\classhint{i}});$.
The field access {\fieldstaticexp {\classhint{1}} {\classhint{2}} {\fld{}} {\e{}}}
calls the field on class {\classhint{1}} with Java signature
${\classhint{2}}\ \fld{};$.
The constructor invocation {\newstaticexp {\overrightarrow{\classhint{i}}} {\classhint{}} 
                               {\class{}} {\overrightarrow{\e{i}}}}
calls the constructor with Java signature
${\classhint{}}\ ({\overrightarrow{\classhint{i}}});$.

%\begin{figure}
%  \footnotesize
%  \begin{mathpar}
%%    \TALocal{}
%%
%%    \TANil{}
%%
%%    \TANewStatic{}
%%
%%    \TALetHint{}
%
%%    \TALet{}
%  \end{mathpar}
%\caption{Type Hint Inference (select rules, figure~\ref{appendix:figure:hintinfer} for full rules)}
%\label{main:figure:hintinfer}
%\end{figure}

We model Clojure's reflection resolution algorithm as a 
rewrite relation
\rewrite {\taenv{}} {\e{}} {\ep{}}
which rewrites \e{} to a possibly-less reflective expression
\ep{} with respect to type hint environment
\taenv{} and Java class table \ct{}.
%
For example,
R-FieldElimRefl emits a non-reflective field if it can
find a field matching the type hint inferred on \ep{}.
  $$
    \RFieldElimRefl{}
  $$

Type hint inference is given by the judgment
\tajudgement {\taenv{}} {\hastype {\e{}} {\tatype{}}}
which infers the (possibly-unknown) type hint \tatype{} of expression \e{} in type hint environment \taenv{}.
We defer the remainder of the rules for rewriting and the definition of type hint inference
to the supplemental material.
%$$
%\TALetHint{}
%$$

As an example, we rewrite
a simple field reference, with the assumptions
that the class \Point{} has a field $x$ of Java type \Number
and \intaenv {\taenv{}} {p} {\Point{}}.
In rewriting
the expression \fieldexp{x}{p},
{\Point{}} is inferred as the type hint of $p$
and \Number{} is the field type
by \fieldtypeliteral{} (\figref{main:figure:javatyping}).
Now we just plug in the new information into our new expression
$$
\fieldstaticexp{\Point}{\Number}{x}{p}.
$$

Now we present the typing rules for resolved Java interoperability.
T-FieldStatic checks a resolved field expression by ensuring the target has
the correct static type, then returns a nilable type corresponding the Java type.

\begin{mathpar}
    {\TFieldStatic}
\end{mathpar}

To continue our example, assume \propenv{} = {\isprop{\Point}{p}}.
T-FieldStatic therefore produces the type
{\Union{\Nil}{\Number}}
for the entire expression.

The rules T-MethodStatic and T-NewStatic work similarly (\figref{main:figure:javatyping}), varying
in the choice of nilability in the conversion function---methods
can return nil but constructors cannot.

>>>>

Recall \egref{example:doexception}.
\begin{minted}{clojure}
...
  (do (if (number? x) nil (throw (new Exception)))
      (inc x)) 
...
\end{minted}

@@

\subsection{Java Interoperability}

\begin{figure}[t]
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \reflectiveexpsyntax{} &\mbox{Expressions}\\
    \\

@@

    \classtableallsyntax{}
  \end{altgrammar}
  $$
  \begin{mathpar}
    {\TNew}

    {\TMethod}

@@

    \BMethod{}
  \end{mathpar}
  \caption{Java Interoperability Syntax, Typing and Operational Semantics}
  \label{main:figure:javatyping}
\end{figure}

@@

We extend the syntax in \figref{main:figure:javatyping} with Java field lookups and calls to
methods and constructors. 
We model the syntax after the `dot' special
form to prevent ambiguity---\clj{(.fld e)} is now \fieldexp{\fld{}}{\e{}},
\clj{(.mth e es*)} is $\methodexp{\mth{}}{\e{}}{\overrightarrow{es}}$
and \clj{(new class es*)} is $\newexp{\class{}}{\overrightarrow{es}}$.

T-Field checks a resolved field expression by ensuring the target has
the correct static type, then returns a nilable type corresponding the Java type.

\begin{mathpar}
    {\TField}
\end{mathpar}

The rules T-Method and T-New work similarly (\figref{main:figure:javatyping}), varying
in the choice of nilability in the conversion function---methods
can return nil but constructors cannot. Void also does not have a constructor.

### 2015-05-16 overview.tex
JVM ops are dynamically typed

Unlike Java, Clojure is dynamically typed. We have no type information about \clj{f}
but we still need to pick a method to call. The Clojure compiler delegates
the choice to runtime using \emph{Java reflection}.
Unfortunately reflection is slow and unpredictable, so Clojure supports \emph{type hints}
to help eliminate it where possible,

\begin{minted}{clj}
  (fn [^File f] (.getParent f))
\end{minted}

Symbols support \emph{metadata}---the syntax \clj{^File f}
is a single expression that is a symbol \clj{f} with metadata \clj{{:tag File}}.
In binding positions like \clj{(fn [^File f] ...)}
syntactic occurrences preserve metadata.

The Clojure compiler uses the type hint
to statically resolve 
the method call to the
\mintinline{java}{public String getParent()}
method of \clj{java.io.File}.
The call to \clj{getParent} is unambiguous at runtime but type 
checking fails---Typed Clojure
considers \clj{f} to be of type \clj{Any}, which is unsafe 
to use as the target of even a resolved method.
If instead we annotate the function parameter with type \clj{File},
we get a static type error.

\begin{minted}{clj}
  (fn [f :- File] (.getParent f)) ;; type error
\end{minted}

Typed Clojure disallows reflection in typed code
so we must add back the type hint to obtain a well-typed
expression.
\newpage
\begin{exmp}
\inputminted[firstline=10,lastline=10]{clojure}{code/demo/src/demo/parent3.clj}
\end{exmp}
A type hint does not correspond to a static property so they are ignored by Typed Clojure.
Type hints are still required, however, as Typed Clojure has no mechanism to provide
optimisations (discussed in \secref{sec:experience}).

@@

We can use {\clj{defmethod}} to install a method to \clj{path}.
\inputminted[firstline=7,lastline=7]{clojure}{code/demo/src/demo/rep.clj}
Now the dispatch table maps
the \emph{dispatch value} \clj{String} to the function
\clj{(fn [x] x)}. 
We add another method
which maps
\clj{File} to the function
\clj{(fn [^File x] (.getPath x))}
in the dispatch table.
\inputminted[firstline=8,lastline=8]{clojure}{code/demo/src/demo/rep.clj}

After installing both methods, the call 
$$
\clj{(path (File. "dir/a"))}
$$
dispatches to the second method we installed because
$$
\clj{(isa? (class "dir/a") String)}
$$
is true, and finally returns 
$$
\clj{((fn [^File x] (.getPath x)) "dir/a")}.
$$

@@

\begin{figure}
\begin{exmp}
\inputminted[firstline=6,lastline=22]{clojure}{code/demo/src/demo/eg7.clj}
\label{example:multidispatch}
\end{exmp}
\begin{exmp}
\inputminted[firstline=6,lastline=20]{clojure}{code/demo/src/demo/eg8.clj}
\label{example:final}
\end{exmp}
\caption{Multimethod Examples}
\end{figure}

>>>>

@@

We can use {\clj{defmethod}} to install a method to \clj{path}.
\inputminted[firstline=7,lastline=7]{clojure}{code/demo/src/demo/rep.clj}
Now the dispatch table maps
the \emph{dispatch value} \clj{String} to the function
\clj{(fn [x] x)}. 
We add another method
which maps
\clj{File} to the function
\clj{(fn [x] (.getPath x))}
in the dispatch table.
\inputminted[firstline=8,lastline=8]{clojure}{code/demo/src/demo/rep.clj}

After installing both methods, the call 
$$
\clj{(path (new File "dir/a"))}
$$
dispatches to the second method we installed because
$$
\clj{(isa? (class "dir/a") String)}
$$
is true, and finally returns 
$$
\clj{((fn [x] (.getPath x)) "dir/a")}.
$$

@@

\begin{figure}
\begin{exmp}
\inputminted[firstline=6,lastline=23]{clojure}{code/demo/src/demo/eg7.clj}
\label{example:multidispatch}
\end{exmp}
\begin{exmp}
\inputminted[firstline=6,lastline=20]{clojure}{code/demo/src/demo/eg8.clj}
\label{example:final}
\end{exmp}
\caption{Multimethod Examples}
\end{figure}
